Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
tracer

\newcommand{\PIR}{\textsf{PIR}\xspace}
\newcommand\pirI[1]{\mathtt{#1}}
\renewcommand{\c}[1]{{\lstinline[style=Rin]!#1!}\xspace}
\newcommand{\code}[1]{{\lstinline[style=Rin]!#1!}\xspace}
% Macros for type names from the old paper.
\newcommand{\attr}[2]{\ensuremath{#1_{\mathtt{#2}}}\xspace}
\newcommand{\attrclass}[3]{\ensuremath{#1^{\mathtt{#3}}_{\mathtt{#2}}}\xspace}
\renewcommand{\to}{\ensuremath{\rightarrow}\xspace}
\newcommand{\D}{\ensuremath{\small\vec{\mathtt D}}\xspace} % Double
\newcommand{\I}{\ensuremath{\small\vec{\mathtt I}}\xspace} % Integer
\renewcommand{\C}{\ensuremath{\small\vec{\mathtt C}}\xspace} % Character
\renewcommand{\L}{\ensuremath{\small\vec{\mathtt L}}\xspace} % Logical
\newcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\X}{\ensuremath{\small\vec{\mathtt X}}\xspace} % Complex
\newcommand{\Y}{\ensuremath{\small\vec{\mathtt Y}}\xspace} % Symbol
\newcommand{\sY}{\ensuremath{\small{\mathtt Y}}\xspace} % Symbol
\newcommand{\sS}{\ensuremath{\small{\mathtt S}}\xspace} % S4
\newcommand{\sF}{\ensuremath{\small{\mathtt F}}\xspace} % Closure
\newcommand{\sE}{\ensuremath{\small{\mathtt E}}\xspace} % Env
\renewcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\sN}{\ensuremath{\small{\mathtt N}}\xspace}     % Null
%\renewcommand{\l}{\ensuremath{\small L<?>}\xspace}     % List
\renewcommand{\l}{\ensuremath{\small\underline{\mathtt ?}}\xspace}     % List
\newcommand{\sD}{\ensuremath{\small{\mathtt D}}\xspace} % Double
\newcommand{\sI}{\ensuremath{\small{\mathtt I}}\xspace} % Integer
\newcommand{\sC}{\ensuremath{\small{\mathtt C}}\xspace} % Character
\newcommand{\sL}{\ensuremath{\small{\mathtt L}}\xspace} % Logical
\newcommand{\sX}{\ensuremath{\small{\mathtt X}}\xspace} % Complex
\newcommand{\sR}{\ensuremath{\small{\mathtt R}}\xspace} % Raw
\newcommand{\ANY}{\ensuremath{\small{\mathtt ?}}\xspace}     % Any
\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt{#1}}}\xspace}     % list<T>
\newcommand{\M}[1]{\ensuremath{\attr{\vec{\tt #1}}{mat}}\xspace}     % matrix
\newcommand{\df}{\ensuremath{\attr{\l}{df}}\xspace}     % data.frame

\newcommand{\contractr}{\emph{contractr}\xspace} % contractr
\newcommand{\roxygen}{\emph{roxygen2}\xspace} % roxygen
\newcommand{\typetracer}{\emph{typetracer}\xspace} % typetracer
\newcommand{\rdt}{\emph{R-dyntrace}\xspace}
\newcommand{\covr}{\emph{covr}\xspace}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Design A Type Language for R}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

\paragraph{Classes} 
R has more than one notion of type. Values can be attributed, one important
attribute is the class of a value.  A class is a list of names that are used
to mimic object-oriented programming. The type system includes class types
writte \CLASS{\ID_1,\dots,\ID_n} A value can have multiple classes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Project Corpus}\label{sec:corpus} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[!h]  \centering
  \includegraphics[width=.8\linewidth]{plots/corpus.pdf}
  \caption{Corpus overview}\label{fig:corpus}
\end{figure}

%
%
%
\section{Methodology}

In this section, we will describe our approach to data collection and
analysis through \typetracer, a dynamic analysis built in the \R-dyntrace
framework.  We will then describe the process by which we distill the rich
information we collect with \typetracer to more digestible ``types'', though
we discuss the precise details of our type annotation language in
Section~\ref{sec:types}.  Finally, we discuss \contractr, a package built in
R which supports our type language and can generate and check contracts for
function types.

%
\subsection{R-dyntrace}
R-dyntrace is a framework for performing dynamic analysis of R. It was designed
and used previously for a large-scale study of laziness in R~\cite{oopsla19} It
is an instrumented R Virtual Machine based on GNU-R version 3.5.0. It exposes
hooks from key points inside the R interpreter to which user defined callbacks
can be attached for gathering information about the program state at those
points. This information can be used to generate detailed program traces for
online or offline analysis. R-dyntrace provides hooks for function entry and
exit, S3/S4 dispatch, non-local jumps (because of longjumps in the R interpreter
on exceptions and other non-local returns), promise creation and forcing,
variable definition and mutation and garbage collection. garbage collection
entry and exit, function entry and exit.

%
%
\subsection{typetracer and Initial Analysis}

To help inform the design of our type system, and to ensure that it aligns
with the day-to-day usage of R, we performed a corpus analysis of some of
the most widely used R packages.  The corpus is discussed in detail in
Section~\ref{sec:corpus}.  We built \typetracer, our dynamic analysis, in
the R-dyntrace~\cite{oopsla19} dynamic analysis framework which is itself
built on a modified version of R targeting R version 3.5.0---we refer the
reader to Section~\ref{sec:r-dyntrace} for more details on the framework.

%
%
\subsubsection{Overview}
The goal of our analysis is to discover argument types for functions,
builtins, and specials (R functions such as \{, \code{if}, etc.).  First,
recall that R wraps arguments in promises (Section~\ref{sec:R}).  Naively
intercepting function calls and recording argument types would result in
incomplete data: While we can dig into promises without forcing them, if the
innermost value is an unevaluated expression we are unable to probe it
without evaluating it, which may alter program behaviour.  To circumvent
this obstacle, we instead intercept the forcing of promises: when a promise
is forced, we identify the surrounding call, collect the type of the value,
and add it to the call trace associated with the current function call.
Once function execution completes, we collect information about the return
value, add it to the call trace, and save the completed call trace to be
written out later.
%
%
\subsubsection{Type Information}
Once a value is obtained, either on function return or when an argument
promise is forced, we can probe it without interfering with program behaviour.
We query the value for the following information:

\begin{itemize}
\item the value's type tag according to R's runtime. R's runtime implements
  type tags as a typedef, e.g. \code{INTSXP} for integers, and, confusingly,
  \code{VECSXP} for lists;
\item its class, which is a list of string class names. Typically, the class
  of a value is held in an attribute with name ``class'', though some
  classes are blessed by the R runtime, such as matrix and array classes;
\item its attributes, a list of metadata attached to the value.
\end{itemize}

Now, depending on the value's type tag, we collect further information:

\begin{itemize}
\item For R's primitive types, we collect its dimension, and if it has a
  names attribute we collect the names as well;
\item For lists, we recursively collect the types of list elements;
\item For lists {\it with a names attribute}, we collect the names as well
  as the types of the slots that those names refer to (data.frames fall into
  this category);
\item For promises, we carefully dig into the promise value {\it without
  forcing it}, and collect information about the innermost value. If the
  innermost value is the unbound value, then the promise is an as-of-yet
  unevaluated expression. In this case, we do our best to guess the value of
  the unevaluated expression: if the expression is not a language
  expression, we collect information about it, otherwise we can say nothing
  without evaluating it and ascribe it type ``any''. \footnote{\AT{I think
      this is confusing, reword.}}
\end{itemize}

@@

Now that we've seen the high-level picture of our approach, we will elaborate on some key details and discuss particular challenges that we faced.

@@

%
%
%
%
\subsection{Post-Processing and Type Ascription}

@@

%
%
%
\subsection{contractr}
\label{sec:contractr}

@@

>>>>

\newcommand{\PIR}{\textsf{PIR}\xspace}
\newcommand\pirI[1]{\mathtt{#1}}
\renewcommand{\c}[1]{{\lstinline[style=Rin]!#1!}\xspace}
\newcommand{\code}[1]{{\lstinline[style=Rin]!#1!}\xspace}
% Macros for type names from the old paper.
\newcommand{\attr}[2]{\ensuremath{#1_{\mathtt{#2}}}\xspace}
\newcommand{\attrclass}[3]{\ensuremath{#1^{\mathtt{#3}}_{\mathtt{#2}}}\xspace}
\renewcommand{\to}{\ensuremath{\rightarrow}\xspace}

\newcommand{\contractr}{{\sf ContractR}\xspace} % contractr
\newcommand{\roxygen}{{\sf Roxygen2}\xspace} % roxygen
\newcommand{\typetracer}{{\sf Typetracer}\xspace} % typetracer
\newcommand{\rdt}{{\sf R-dyntrace}\xspace}
\newcommand{\covr}{{\sf Covr}\xspace}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Designing A Type Language for R}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\paragraph{Classes} 
R has more than one notion of type. Values can be attributed, one important
attribute is the class of a value.  A class is a list of names that are used
to mimic object-oriented programming. The type system includes class types
written \CLASS{\ID_1,\dots,\ID_n}.

@@

%
%
\section{Extracting and Checking Signatures}

For this paper, we have built tooling to (\emph{a}) automate the extraction
of raw type signatures from execution traces, (\emph{b}) infer type
signatures from a set of raw types, and (\emph{c}) validate the inferred
signatures by the means of contracts. This section presents each step
of this pipeline.

\subsection{Extracting raw type signature from traces}

We implemented \typetracer an automated tool for extracting raw type
signatures from execution traces of R programs. The goal of this tool is to
output a tuple $\langle f, t_1, \dots, t_n, t\rangle$ for each function call
during the execution of a program, where $f$ is an identifier for a
function, $t_i$ are type-level summaries of the arguments and $t$ is a
summary of the return value.

While this task is seemingly simple, the details and their proverbial devil
are surprisingly tricky to get right and to scale to long running programs.
To avoid starting from scratch, our implementation reuses an open source
dynamic analysis framework for R named \rdt~\cite{oopsla19} which consists
of an instrumented R Virtual Machine based on GNU-R version 3.5.0. The
framework exposes hooks at key points in the R interpreter to which user
defined callbacks can be attached. These hooks include function entry and
exit, method dispatch for the S3 and S4 object systems, the longjumps used
by the interpreter to implement non-local exit, creation and forcing of
promises (lazy evaluation), variable definition, value creation, mutation
and garbage collection.

\paragraph{Raw types}
The type information output by the tool includes the \emph{type tag} of each
value. The R internal types are translated to the type name of the proposed
type system. The next bit of information output for each value is its
\emph{class}.  The class is an optional list of names, it may be absent,
and, in some cases, it may be implicit (i.e. the interpreter blesses some
values with the \k{matrix} and \k{array} classes even when they have no
attribute). Finally, for each value, the tool outputs the list of
\emph{attribute names} (other than \k{class}).

Depending on the value's type tag, the tool collects further information:

\begin{itemize}
\item For vectors, the number of dimensions and their names are output;
\item For lists, a recursive traversal collects element types and
  the name attributes of elements.  
  TODO: HOW DO WE OUTPUT HETEROGENEOUS NAMES???
\item For promises, we carefully dig into the promise value {\it without
  forcing it}, and collect information about the innermost value. If the
  innermost value is the unbound value, then the promise is an as-of-yet
  unevaluated expression. In this case, we do our best to guess the value of
  the unevaluated expression: if the expression is not a language
  expression, we collect information about it, otherwise we can say nothing
  without evaluating it and ascribe it type ``any''. 
\end{itemize}

@@

The goal of our analysis is to discover argument types for functions,
builtins, and specials (R functions such as \{, \code{if}, etc.).  First,
recall that R wraps arguments in promises (Section~\ref{sec:R}).  Naively
intercepting function calls and recording argument types would result in
incomplete data: While we can dig into promises without forcing them, if the
innermost value is an unevaluated expression we are unable to probe it
without evaluating it, which may alter program behaviour.  To circumvent
this obstacle, we instead intercept the forcing of promises: when a promise
is forced, we identify the surrounding call, collect the type of the value,
and add it to the call trace associated with the current function call.
Once function execution completes, we collect information about the return
value, add it to the call trace, and save the completed call trace to be
written out later.

@@

\subsection{Inferring type signatures from raw types}

@@

%
%
%
\subsection{Checking types signatures with contracts}

@@

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Project Corpus}\label{sec:corpus} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[!h]  \centering
  \includegraphics[width=.98\linewidth]{plots/corpus.pdf}
  \caption{Corpus overview}\label{fig:corpus}
\end{figure}
